### 2020-07-12 thesis/thesis.tex
[thesis] Break some lines
[thesis] My insights about computational effects

%%%%% WLASNE DODATKOWE PAKIETY
% \usepackage{graphicx,listings,tikz}
\usepackage{syntax}
\usepackage{mathalfa}

\usepackage[backend=bibtex]{biblatex}
\addbibresource{mybib.bib}
%
%%%%% WŁASNE DEFINICJE I POLECENIA
%
%\theoremstyle{definition} \newtheorem{definition}{Definition}[chapter]
%\theoremstyle{remark} \newtheorem{remark}[definition]{Observation}
%\theoremstyle{plain} \newtheorem{theorem}[definition]{Theorem}
%\theoremstyle{plain} \newtheorem{lemma}[definition]{Lemma}
%\renewcommand \qedsymbol {\ensuremath{\square}}
% ...
%%%%%

@@

%%%%% POCZĄTEK ZASADNICZEGO TEKSTU PRACY

@@

% TODO This requires a rework
In this thesis we present experimental programming language Freak, which is an
implementation of Continuation Passing Style for Effect Handlers paper~\cite{handlers-cps},
with the addition of a few basic constructs. We start by presenting the related work,
then discuss syntax and operational semantics. Basic usage guide for playing with
the language is provided, as well as implementation details and examples. We conclude
by stating what are the possible augmentations, that are intended to be made in the future.

@@

\chapter{Background}
\section{Algebraic Effects}

@@

@@

@@

>>>>

%%%%% WLASNE DODATKOWE PAKIETY
% \usepackage{graphicx,listings,tikz}
\usepackage{syntax}
\usepackage{mathalfa}
\usepackage{textcomp}

\usepackage[backend=bibtex]{biblatex}
\addbibresource{mybib.bib}
%
%%%%% WŁASNE DEFINICJE I POLECENIA
%
\theoremstyle{definition} \newtheorem{definition}{Definition}[chapter]
\theoremstyle{remark} \newtheorem{remark}[definition]{Observation}
\theoremstyle{plain} \newtheorem{theorem}[definition]{Theorem}
\theoremstyle{plain} \newtheorem{lemma}[definition]{Lemma}
%\renewcommand \qedsymbol {\ensuremath{\square}}
% ...
%%%%%

@@

@@

\textit{
    My algebraic methods are really methods of working and thinking; this is why
    they have crept in everywhere anonymously. $\sim{}$Emmy Noether
}

% TODO This requires a rework

In this thesis we present experimental programming language Freak, which is an
implementation of Continuation Passing Style for Effect Handlers paper~\cite{handlers-cps},
with the addition of a few basic constructs. We start by presenting the related work,
then discuss syntax and operational semantics. Basic usage guide for playing with the
language is provided, as well as implementation details and examples. We conclude by
stating what are the possible augmentations, that are intended to be made in the future.

@@

\chapter{Background}

\section{Computational Effects}

Since the rapid development of computational theory in 1930s by A. Turing, K. Godel and
A. Church, we have a well-established notion of what can and what cannot be done
through algorithmic means, which we can almost directly translate to being computable
by our machines. Through next years we have developed mainstream languages that are used
almost everywhere, with a great success.

Under these circumstances one may pose a question, why do we still bother with
development of languages theory, since so much has been done already. Is there
anything that drives us towards further research? Indeed, one active branch
revolves around equational theory to asses equality of two programs, which we
know that in general setting is undecidable. Proof methods may include extensional,
contextual or logical equality. However, there is no doubt that these formal
ways of reasoning about programs, while being crucial for assessing correctness,
do not bring direct benefits for practical, everyday use cases.

% Not sure I like this part. Isn't it too broad to say handle complexity?
Other branch of languages theory, that we shall investigate more in this thesis,
is about handling complexity of programs. Various methods of static analysis has
been developed, most notably, type systems.

What is the source of complexity? Thanks to strong and static type systems along
with their implementations, we have solid tools to work efficiently on functions
that are pure. That being said, we claim that the sole complexity of programs
comes from side effects, which we cannot avoid in writing anything useful.

We need to have a good way for handling computational effects, which so far has
been modeled through monads~\cite{moggi}.

However, they were found to be, to say the least, cumbersome to work with when
complexity and number of different effects increases. It is perhaps not a coincidence
that many functional programming languages do not have them, because of the
non-composable nature of them.

Having functions $ f: a \rightarrow m b $ and $ g : b \rightarrow m' c $, it
should not be an issue to compose them together.

Recall that the main idea of monads lies on the fact that we provide a function
to lift value into a functor and associative kleisli composition operator to combine
computations together. It turns out, that it becomes complex and unpleasent to
combine two functors together to form a new monad. For this purpose, monad
transformers arose in Haskell.

Most of the common languages avoid this by not expressing computational effects
in the type system, and instead one may think about functions as being implicitly
embed in a Kleisli Category over a functor T, where T is a hidden signature over
all possible side effects that occur in our program.

Not only we would like to bring back effects to our type system, but also do it
in a way that is composable. This is where algebraic effects comes to the rescue.

From theoretical point of view, we need to develop equa about our effects
to assert correctness of our langauge as well as to have the right hammer
to reason about our programs.

That is the point where we would like to introduce to unfamiliar read a notion
of algebraic effects.

\section{Algebraic Effects}

@@

% Algebraic effects as Interfaces for side effects

Algebraic effects can be thought as an public interface for computational effects.
Declarative approach allow us to write programs in which the actual semantic of
source code is dependent on handler that defines the meaning of a subset of effects.

This is really a breakthrough from practical point of view, as we may substitute
logic depending on execution environment. As an example, fetching for resources
can behave differently as we run tests, debug our code, or run it on production.
In the same manner, they neatly allow us to abstract over implementation details.

@@

    \subsection{Coinductive Reasoning}

    Coinduction, and coinductive structures, provide us a way of observation
    of the behaviour, as a contractry to construction in inductive reasoning.
    Simple example of the duality can be expressed through induction over finite
    list and coinductive observation of infinite streams.

    Doing the latter may involve modification of the internal state of the machine
    that is generating the infinite streams, or in more concrete scenario,
    alternation of the external resource that is providing us the data.

    This is one of the cases where interaction with external resource multiple
    times, or more specifically in case of algebraic effects, invoking resumption more
    than once, may lead to unexpected behaviour that would not be expected in standard
    control flow. This captures the excessive generality of effects, and is the
    issue that we would like to address with coalgebraic effects.

    % TODO: On it's own this snippet a bit dry and of course requires additional examples,
    % for instance the classic one with operating on closed file

@@

Detecting whether resumption is called only once is undecidable.
